UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2018-11-16 ms/main.tex
shorter abstract

\begin{abstract}
%Max 250
Background: Genetic structure in large cohorts can arise from technical, sampling and demographic variation. Data visualisation is therefore a first step in most genomic analyses. However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure in large cohorts. We investigate an approach to dimension reduction and visualization of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
\\
\\
Results: Using genotype data from the 1000 Genomes Project, the Health and Retirement Study, and the UK Biobank, we demonstrate that PCA-UMAP effectively clusters individuals who are genetically closely related while placing them in a global continuum of genetic variation. This dimension reduction approach reveals subpopulations within the American Hispanic population, fine-scale relationships between geography and genotype in the UK population, and rich phenotypic variation across space and ethnicity in the UK Biobank. 
\\
\\
Conclusion: PCA-UMAP reveals genetic subgroups among study participants that had been previously overlooked, opening new lines of investigation for demographic research and statistical genetics. Given its small computational cost, it provides an important general-purpose approach to exploratory analysis in population-scale datasets.
\end{abstract}

>>>>

\begin{abstract}
%Max 250
Genetic structure in large cohorts results from technical, sampling and demographic variation. Visualisation is therefore a first step in most genomic analyses. However, existing data exploration methods struggle with unbalanced sampling and the many scales of population structure. We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to reveal layers of diversity and cryptic population structure.
Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP effectively clusters closely related individuals while placing them in a global continuum of genetic variation. This approach reveals previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population. 
This opens new lines of investigation for demographic research and statistical genetics. Given its small computational cost, PCA-UMAP also provides a general-purpose approach to exploratory analysis in population-scale datasets.
\end{abstract}
